---
title: "Krug Skin Only - DSP Analysis QC Report"
format:
  html:
    code-fold: true
editor: visual
#css: "qc_report_style.css"

params:
  dsp.analysis.folder: "/Users/cauleyes/CPTR/DSP_Analysis/"
  data.folder: "/Users/cauleyes/CPTR/CPTR-5_Krug_Ramaswami/data/"
  results.folder: "/Users/cauleyes/CPTR/CPTR-5_Krug_Ramaswami/results/"
  project.folder: "/Users/cauleyes/CPTR/CPTR-5_Krug_Ramaswami/"
  run.folder: "qc_report/"
  dataset.name: "Krug_skin_only"
---

## Load Libraries

```{r Load Libraries}
#| warning: false
#| message: false

# Load all relevant libraries

library(DSPWorkflow)
library(GeomxTools)
library(dplyr)
library(limma)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(ggforce)
library(shadowtext)
library(stringr)
library(PCAtools)
library(readxl)
library(gridExtra)
library(grid)
library(knitr)
library(gt)
library(tidyr)
library(openxlsx)
library(ComplexUpset)
library(reshape2)
library(cowplot)
library(preprocessCore)

source(paste0(params$dsp.analysis.folder, "DSP_QC_functions.R"))
```

## Initialization

```{r Initialization}

# Input file parameters

pkc.file.name <- "Hs_R_NGS_WTA_v1.0.pkc"
pkc.file.path <- paste0(params$data.folder, pkc.file.name)

annotation.file.name <- "annotation_Krug_CPTR_5_October_2024_NC_edit_SK_Skin_metadata.xlsx"
annotation.file.path <- paste0(params$project.folder, "NIDAP_Workbook_Restoration/", annotation.file.name)

dcc.files.skin <- list.files(file.path(paste0(params$data.folder, "dcc_skin_subset")),
  pattern = ".dcc$",
  full.names = TRUE,
  recursive = TRUE
)

dcc.files <- list.files(file.path(paste0(params$data.folder, "dcc")),
  pattern = ".dcc$",
  full.names = TRUE,
  recursive = TRUE
)

# Annotation parameters
annotation.sheet.name <- "annotation"
sample.id.field.name <- "Sample_ID"
roi.field.name <- "ROI"
panel.field.name <- "panel"
class.field.name <- "class"
region.field.name <- "Region"
segment.field.name <- "segment"
area.field.name <- "area"
nuclei.field.name <- "nuclei"
exclude.sankey <- FALSE
segment.id.length <- 8

# Create the GeoMxSet Object

init.object <- initialize_object(dcc.files = dcc.files,
                            pkc.files = pkc.file.path,
                            annotation.file = annotation.file.path,
                            annotation.sheet.name = annotation.sheet.name,
                            sample.id.field.name = sample.id.field.name,
                            roi.field.name = roi.field.name,
                            panel.field.name = panel.field.name, 
                            class.field.name = class.field.name, 
                            region.field.name = region.field.name, 
                            segment.field.name = segment.field.name,
                            area.field.name = area.field.name,
                            nuclei.field.name = nuclei.field.name, 
                            segment.id.length = segment.id.length)

```

## Object Summary

@fig-sankey shows a summary of AOIs per annotation

```{r Object Summary, fig.width=12, fig.height=8}
#| label: fig-sankey
#| fig-cap: "Sankey Plot"
#| warning: false

sankey.output <- plot_sankey(object = init.object, 
                      lane.1 = "slide_name", 
                      lane.1.order = c("_1B", "_2", "_3", "_4"), 
                      lane.2 = "class", 
                      lane.2.order = c("Biopsy_Skin", "PDX_Skin"), 
                      lane.3 = "region", 
                      lane.3.order = c("tumor", "vessel"), 
                      lane.4 = "segment", 
                      lane.4.order = c("LANA_pos", "full_ROI"), 
                      fill.lane = "region")

print(sankey.output$sankey.plot)
```

@fig-aoibarplot shows the total AOI counts per annotation

```{r AOI Count Bar Plot, fig.width=12, fig.height=8}
#| label: fig-aoibarplot
#| fig-cap: "AOI Count Bar Plot"
#| warning: false

print(sankey.output$AOI.bar.plot)
```

## QC and Filtering

```{r QC and Filtering}


qc.output <-  qcProc(object = init.object,
                        min.segment.reads = 1000, 
                        percent.trimmed = 80,    
                        percent.stitched = 75,   
                        percent.aligned = 80,    
                        percent.saturation = 50, 
                        min.negative.count = 1,   
                        max.ntc.count = 9000,     
                        min.nuclei = 20,         
                        min.area = 1000,
                        print.plots = FALSE, 
                     shift.counts.zero = FALSE, 
                     remove.local.outliers = TRUE
                     )
    

```

Summary of QC for AOIs and Probes

```{r QC Summary}

qc.output$table

```

#### AOI QC

AOI distribution by parameter and annotation

```{r AOI Plots}

# Print AOI plots
qc.output$plot$trimmed
qc.output$plot$aligned
qc.output$plot$stitched
qc.output$plot$saturated
qc.output$plot$neg.plot


```

AOIs that have been flagged with the given QC parameters

aoi.flag.table \<- aoi_flag_table(aoi.flags = qc.output\$segment.flags)

print(aoi.flag.table)

```{r AOI Flags}

aoi.flag.table <- aoi_flag_table(aoi.flags = qc.output$segment.flags)

aoi.flag.table

```

### Probe QC

Probes that have been flagged as either local or global outliers.

```{r Probe Flag Tables}


probe.flag.output <- probe_flag_table(probe.flags = qc.output$probe.flags, 
                                      object = qc.output$object)

probe.flag.output$probe.flag.summary

probe.flag.output$probe.flags.table


```

### Filtering

```{r Filtering}

# Add LOQ and calculate detection of genes per AOI
loq.detection.output <- loq_detection(object = qc.output$object, 
              pkc.file.names = pkc.file.name)
```

#### Overall Gene Detection per AOI

@fig-GeneDetectionbByAOI shows detection rate per AOI, colored by region.

```{r Overall AOI Detection}
#| label: fig-AOIDetection
#| tbl-cap: "AOI Detection"
#| warning: false

aoi.detection <- aoi_detection(object = loq.detection.output$object)

print(aoi.detection$detection.bar.plot)

```

AOIs in the low detection bin of 1-5%

```{r Low Detection AOI}
#| warning: false
#| message: false

aoi.detection$low.detection.table

```

```{r Gene Detection All AOIs}
#| warning: false
#| message: false

# Store the AOI detection for export in the QC summary
aoi.detection.summary <- pData(loq.detection.output$object) %>% 
  select(any_of(c("segmentID", "GeneDetectionRate", "DetectionThreshold")))

```

##### Filter out AOIs with low detection

```{r Filter by AOI}
#| warning: false
#| message: false

# Filter the data using the cutoff for gene detection rate
aoi.gene.rate.cutoff <- 10

object <- loq.detection.output$object

object.aoi.filtered <-
    object[, pData(object)$GeneDetectionRate >= aoi.gene.rate.cutoff]

# Create a table of AOIs that are filtered out
aoi.filtered.out <- pData(object[, pData(object)$GeneDetectionRate < aoi.gene.rate.cutoff])

aoi.filtered.out.table <- aoi.filtered.out %>% 
  select(segmentID, GeneDetectionRate) %>% 
  gt() %>% 
  tab_header(title = "AOIs Filtered Out", 
             subtitle = paste0(aoi.gene.rate.cutoff, "% Cutoff"), preheader = NULL)

aoi.filtered.out.table

```

#### Detection per Gene

```{r Detection per Gene}
#| warning: false
#| message: false

# Rerun LOQ to calculate new detection rate for individual genes:
loq.detection.output.aoi.filtered <- loq_detection(object = object.aoi.filtered, 
                                      pkc.file.names = pkc.file.name)
```

@fig-GeneDetection shows the gene detection as a percent of all AOIs

```{r Gene Detection Plot}
#| label: fig-GeneDetection
#| fig-cap: "Gene Detection Percent of All AOIs"
#| warning: false

gene.detection <- gene_detection(object = loq.detection.output.aoi.filtered$object, 
                                    facet.column = "region", 
                                    loq.mat = loq.detection.output.aoi.filtered$loq.matrix)

print(gene.detection$total.plot)
```

@fig-FacetGeneDetection shows the gene detection as a percent of all AOIs per annotation group

```{r Facet Gene Detection Plot}
#| label: fig-FacetGeneDetection
#| fig-cap: "Gene Detection Percent of All AOIs per Annotation Group"
#| warning: false

print(gene.detection$facet.plot)

```

Gene detection rates for specified genes of interest

```{r Genes of Interest}
#| label: tbl-GenesOfInterest
#| tbl-cap: "Gene of Interest Detection Rate"
#| warning: false

# Gene of interest detection table
goi <- c("A2M", "CD44")

goi.df <- data.frame(Gene = goi,
                     Number = fData(loq.detection.output.aoi.filtered$object)[goi, "DetectedSegments"],
                     DetectionRate = fData(loq.detection.output.aoi.filtered$object)[goi, "DetectionRate"])

# Print the GOI table
goi.table <- goi.df %>% gt()

goi.table
```

@fig-DetectionPerGeneLoss shows the loss of percentage of all AOIs individual genes are detected within

```{r Gene Detection Rates Loss Plot}
#| label: fig-DetectionPerGeneLoss
#| fig-cap: "Gene Detection Loss"
#| warning: false

print(gene.detection$detect.loss.plot)

```

Summary of all gene detection rates

```{r Gene Detection Rate Summary}
#| warning: false
#| message: false

# Gather a summary of the every gene's detection percentage in all AOIs
gene.detection.summary <- fData(loq.detection.output.aoi.filtered$object) %>% 
  mutate(gene = rownames(fData(loq.detection.output.aoi.filtered$object))) %>% 
  select(any_of(c("gene", "DetectionRate", "DetectionThreshold")))


```

##### Filter out genes with low detection

```{r Filter Genes}
#| warning: false
#| message: false

# Set the cutoff for gene detection (percentage)
gene.detection.cutoff <- 10

# Manually include the negative control probe, for downstream use
negative.probe.fData <- subset(fData(loq.detection.output.aoi.filtered$object), CodeClass == "Negative")
neg.probes <- unique(negative.probe.fData$TargetName)

# Subset for genes above the study gene detection rate cutoff
object.gene.filtered <- loq.detection.output.aoi.filtered$object[fData(loq.detection.output.aoi.filtered$object)$DetectionRate >= gene.detection.cutoff |
                   fData(loq.detection.output.aoi.filtered$object)$TargetName %in% neg.probes, ]

# Create a summary of genes that are filtered out
genes.filtered.out <- fData(loq.detection.output.aoi.filtered$object[fData(loq.detection.output.aoi.filtered$object)$DetectionRate < gene.detection.cutoff, ])

print(paste0(nrow(genes.filtered.out), " genes were filtered out of the GeoMxSet Object. For the full list, check the QC Output File."))

```

**Write QC Output File**

```{r Write QC output}

write.qc.file <- TRUE

if(write.qc.file == TRUE){
  
  # Start the QC output excel workbook
  qc.info.output <- createWorkbook()
  
  # Add the AOI flag info to the output file
  addWorksheet(qc.info.output, "AOI QC Flags")
  writeData(qc.info.output, sheet = "AOI QC Flags", qc.output$segment.flags)
  
  # Add the probe flag QC info to the output file
  addWorksheet(qc.info.output, "Probe QC Flags")
  writeData(qc.info.output, sheet = "Probe QC Flags", probe.flag.output$probe.flags.table)

  # Add the AOI detection QC info
  addWorksheet(qc.info.output, "AOI Detection Rate")
  writeData(qc.info.output, sheet = "AOI Detection Rate", aoi.detection.summary)
  
  # Add the Gene detection QC info to the output file
  addWorksheet(qc.info.output, "Gene Detection Rate")
  writeData(qc.info.output, sheet = "Gene Detection Rate", gene.detection.summary)
  
  # Save the QC output file
  saveWorkbook(qc.info.output, paste0(params$results.folder, params$run.folder, "QC_info.xlsx"), overwrite = TRUE)
  
}


```

#### Q3 versus Negative Background

```{r Visualize Density of Q3 Scores versus Negative Background, warning=FALSE, message=FALSE}
#| label: fig-densityQ3vsBackground
#| fig-cap: "Density of Q3 Normalized Counts versus Background"
#| warning: false

plot.distribution <- plot_distribution(object = object.gene.filtered, 
                                       facet.annotation = "region")

print(plot.distribution$distr.plots)


```

### Normalization

```{r Normalization, warning=FALSE, message=FALSE}

q3.normalization.output <- geomxNorm(
                                  object = object.gene.filtered, 
                                  norm = "q3")
    
    
neg.normalization.output <- geomxNorm(
                                  object = object.gene.filtered, 
                                  norm = "neg")


# Quantile normalization
quantile.normalization.counts <- normalize.quantiles(object.gene.filtered@assayData$exprs)

dimnames(quantile.normalization.counts) = dimnames(object.gene.filtered@assayData$exprs)

quantile.normalization.counts <- as.data.frame(quantile.normalization.counts)


# Set up the mapping df for AOI IDs
mapping.df.segmentID <- data.frame(sample_ID = rownames(annotation), 
                         segmentID = annotation$segmentID)


# Q3 normalization clean up
object <- q3.normalization.output$object
q3.counts <- as.data.frame(object@assayData$q_norm)
  
# Add genes as a column and move to the left
q3.counts$gene <- rownames(q3.counts)
q3.counts <- q3.counts %>% 
  select(gene, everything())

# Replace the sampleID column names with human readable AOI ID
colnames(q3.counts)[match(mapping.df.segmentID$sample_ID, colnames(q3.counts))] <- mapping.df.segmentID$segmentID

export.counts <- FALSE

if(export.counts){
  
  write.csv(q3.counts, file = paste0(params$project.folder, 
                                     "NIDAP_Workbook_Restoration/", 
                                     params$dataset.name, "_q3normalized_counts.csv"), 
            row.names = FALSE)
  
}
    

export.q3norm.object <- FALSE 
if(export.q3norm.object == TRUE){
  
  object <- q3.normalization.output$object
  annotation <-pData(object)
  counts <- object@assayData$q_norm
  
  saveRDS(object, 
          file = paste0(params$project.folder, 
                        "NIDAP_Workbook_Restoration/", 
                        params$dataset.name, "_normalized_object.RDS"))
}   


```

#### RLE Plots

```{r RLE Plots}
#| warning: false
#| message: false

# Counts
raw.counts <- q3.normalization.output$object@assayData$exprs
q3.counts <- q3.normalization.output$object@assayData$q_norm
neg.counts <- neg.normalization.output$object@assayData$neg_norm

# Annotation
annotation <- pData(q3.normalization.output$object)

# Annotation to groupo by
annotation.facet <- "slide_name"

# The ammount to subsample, with 1 being no subsampling
subsample.ammount <- 0.8

raw.rle <- make_rle_plot(counts = raw.counts, 
                         annotation = annotation, 
                         annotation.facet = annotation.facet, 
                         subsample.ammount = subsample.ammount)

q3.rle <- make_rle_plot(counts = q3.counts, 
                         annotation = annotation, 
                         annotation.facet = annotation.facet, 
                         subsample.ammount = subsample.ammount)

neg.rle <- make_rle_plot(counts = neg.counts, 
                         annotation = annotation, 
                         annotation.facet = annotation.facet, 
                         subsample.ammount = subsample.ammount)

quant.rle <- make_rle_plot(counts = quantile.normalization.counts, 
                         annotation = annotation, 
                         annotation.facet = annotation.facet, 
                         subsample.ammount = subsample.ammount)


print(raw.rle)
print(q3.rle)
print(neg.rle)
print(quant.rle)



```

#### **Example AOIs**

```{r Normalization Effects on Counts, fig.width=12, fig.height=8}
#| label: fig-NormEffects
#| fig-cap: "Normalization Effects on Counts"
#| warning: false
#| message: false

# The raw counts boxplot
transform1.raw<- exprs(q3.normalization.output$object[,1:30])
transform2.raw<- as.data.frame(transform1.raw)
transform3.raw<- melt(transform2.raw)
ggboxplot.raw <- ggplot(transform3.raw, aes(variable, value)) +
  stat_boxplot(geom = "errorbar") +
  geom_boxplot(fill="grey") +
  scale_y_log10() +
  xlab("Example AOIs") + 
  ylab("Counts, Raw") +
  ggtitle("Neg Norm Counts") +
  scale_x_discrete(labels=c(1:30))

# The Q3 normalized counts boxplot
transform1.norm <- assayDataElement(q3.normalization.output$object[,1:30], elt = "q_norm")
transform2.norm <- as.data.frame(transform1.norm)
transform3.norm <- melt(transform2.norm)
ggboxplot.q3norm <- ggplot(transform3.norm, aes(variable, value)) +
  stat_boxplot(geom = "errorbar") +
  geom_boxplot(fill="cadetblue2") +
  scale_y_log10() +
  xlab("Example AOIs") + 
  ylab("Counts, Q3 Normalized") +
  ggtitle("Q3 Norm Counts") +
  scale_x_discrete(labels=c(1:30))

# The Negative normalized counts boxplot
transform1.norm <- assayDataElement(neg.normalization.output$object[,1:30], elt = "neg_norm")
transform2.norm <- as.data.frame(transform1.norm)
transform3.norm <- melt(transform2.norm)
ggboxplot.negnorm <- ggplot(transform3.norm, aes(variable, value)) +
  stat_boxplot(geom = "errorbar") +
  geom_boxplot(fill="indianred") +
  scale_y_log10() +
  xlab("Example AOIs") + 
  ylab("Counts, Neg. Normalized") +
  ggtitle("Neg Norm Counts") +
  scale_x_discrete(labels=c(1:30))

# The Negative normalized counts boxplot
transform1.norm <- quantile.normalization.counts[,1:30]
transform2.norm <- melt(transform1.norm)
ggboxplot.quantnorm <- ggplot(transform2.norm, aes(variable, value)) +
  stat_boxplot(geom = "errorbar") +
  geom_boxplot(fill="darkseagreen") +
  scale_y_log10() +
  xlab("Example AOIs") + 
  ylab("Counts, Neg. Normalized") +
  ggtitle("Neg Norm Counts") +
  scale_x_discrete(labels=c(1:30))

print(ggboxplot.raw)
print(ggboxplot.q3norm)
print(ggboxplot.negnorm)
print(ggboxplot.quantnorm)


```

#### Principal Component Analysis (PCA)

```{r PCA, warning=FALSE, message=FALSE}

# See reference vignette: https://bioconductor.org/packages/release/bioc/vignettes/PCAtools/inst/doc/PCAtools.html#introduction

# Load the Geomx objects
object.q3 <- q3.normalization.output$object
object.neg <- neg.normalization.output$object

# Gather the the normalized counts
q3.norm.counts.df <- as.data.frame(object.q3@assayData$q_norm)
neg.norm.counts.df <- as.data.frame(object.neg@assayData$neg_norm)
quant.norm.counts.df <- quantile.normalization.counts

# Convert counts to log2
q3.log.counts.df <- q3.norm.counts.df %>% 
  mutate_all(~ log2(.)) %>% 
  rename_all(~ gsub("\\.dcc", "", .))
neg.log.counts.df <- neg.norm.counts.df %>% 
  mutate_all(~ log2(.)) %>% 
  rename_all(~ gsub("\\.dcc", "", .))
quant.log.counts.df <- quant.norm.counts.df %>% 
  mutate_all(~ log2(.)) %>% 
  rename_all(~ gsub("\\.dcc", "", .))


# Remove the negative controls from the log counts
control.probes <- c("NegProbe-WTX")
q3.log.counts.df <- q3.log.counts.df[!(rownames(q3.log.counts.df) %in% control.probes), ]
neg.log.counts.df <- neg.log.counts.df[!(rownames(neg.log.counts.df) %in% control.probes), ]
quant.log.counts.df <- quant.log.counts.df[!(rownames(quant.log.counts.df) %in% control.probes), ]


# Load the annotation (same for both normalization types)
annotation <- pData(object.q3)

# Remove NTCs
cleaned.annotation.df <- as.data.frame(annotation[annotation$'slide_name' != "No Template Control", ])

# Order of rownames of annotation need to match columns of count data
cleaned.annotation.df <- cleaned.annotation.df[order(rownames(cleaned.annotation.df)), ]

q3.log.counts.df <- q3.log.counts.df[order(colnames(q3.log.counts.df))]
neg.log.counts.df <- neg.log.counts.df[order(colnames(neg.log.counts.df))]
quant.log.counts.df <- quant.log.counts.df[order(colnames(quant.log.counts.df))]

# Remove .dcc from Sample ID row names
cleaned.annotation.df <- cleaned.annotation.df %>% `rownames<-`(sub("\\.dcc", "", rownames(.)))

# Generate a PCA table for all samples for both normalization types
q3.pca.table <- pca(q3.log.counts.df, 
                 metadata = cleaned.annotation.df, 
                 removeVar = 0.1)
neg.pca.table <- pca(neg.log.counts.df, 
                 metadata = cleaned.annotation.df, 
                 removeVar = 0.1)
quant.pca.table <- pca(quant.log.counts.df, 
                 metadata = cleaned.annotation.df, 
                 removeVar = 0.1)

```

#### PCA by Segment

```{r PCA for Q3 segment, fig.width=12, fig.height=8}
#| label: fig-PCAsegmentQ3
#| fig-cap: "PCA colored by Segment for Q3 Normalization"
#| warning: false

q3.pca.plot.segment <- biplot(q3.pca.table, 
                         colby = "segment", 
                         legendPosition = "right", 
                         legendLabSize = 6, 
                         legendIconSize = 3, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")

print(q3.pca.plot.segment)

```

```{r PCA for Negative segment, fig.width=12, fig.height=8}
#| label: fig-PCAsegmentNeg
#| fig-cap: "PCA colored by Segment for Negative Normalization"
#| warning: false

neg.pca.plot.segment <- biplot(neg.pca.table, 
                         colby = "segment", 
                         legendPosition = "right", 
                         legendLabSize = 6, 
                         legendIconSize = 3, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.segment)

```

#### PCA by Region

```{r PCA for Q3 region, fig.width=12, fig.height=8}
#| label: fig-PCAregionQ3
#| fig-cap: "PCA colored by Region for Q3 Normalization"
#| warning: false

q3.pca.plot.region <- biplot(q3.pca.table, 
                         colby = "region", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")


print(q3.pca.plot.region)

```

```{r PCA for Neg region, fig.width=12, fig.height=8}
#| label: fig-PCAregionNeg
#| fig-cap: "PCA colored by Region for Negative Normalization"
#| warning: false

neg.pca.plot.region <- biplot(neg.pca.table, 
                         colby = "region", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.region)

```

#### PCA by Class

```{r PCA for Q3 class, fig.width=12, fig.height=8}
#| label: fig-PCAclassQ3
#| fig-cap: "PCA colored by Class for Q3 Normalization"
#| warning: false

q3.pca.plot.class <- biplot(q3.pca.table, 
                         colby = "class", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")

print(q3.pca.plot.class)

```

```{r PCA for Negative class, fig.width=12, fig.height=8}
#| label: fig-PCAclassNeg
#| fig-cap: "PCA colored by Class for Negative Normalization"
#| warning: false

neg.pca.plot.class <- biplot(neg.pca.table, 
                         colby = "class", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.class)

```

#### PCA by Slide

```{r PCA for Q3 slide, fig.width=12, fig.height=8}
#| label: fig-PCAslideQ3
#| fig-cap: "PCA colored by Slide for Q3 Normalization"
#| warning: false


q3.pca.plot.slide <- biplot(q3.pca.table, 
                         colby = "slide_name", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")

print(q3.pca.plot.slide)

```

```{r PCA for Neg slide, fig.width=12, fig.height=8}
#| label: fig-PCAslideNeg
#| fig-cap: "PCA colored by Slide for Negative Normalization"
#| warning: false

neg.pca.plot.slide <- biplot(neg.pca.table, 
                         colby = "slide_name", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.slide)

```

### PCA for Publication

#### Supp Figure 6E

```{r}


object.q3 <- q3.normalization.output$object 

# Gather the the normalized counts
q3.norm.counts.df <- as.data.frame(object.q3@assayData$q_norm)

# Convert counts to log2
q3.log.counts.df <- q3.norm.counts.df %>% mutate_all(~ log2(.)) %>% rename_all(~ gsub("\\.dcc", "", .))

# Remove the negative controls from the log counts
control.probes <- c("NegProbe-WTX")
q3.log.counts.df <- q3.log.counts.df[!(rownames(q3.log.counts.df) %in% control.probes), ]

# Load the annotation (same for both normalization types)
annotation <- pData(object.q3)
rownames(annotation) <- rownames(pData(object.q3))
annotation$sample_ID <- rownames(annotation)

# Remove NTCs
cleaned.annotation.df <- as.data.frame(annotation[annotation$'slide_name' != "No Template Control", ])

# Order of rownames of annotation need to match columns of count data
cleaned.annotation.df <- cleaned.annotation.df[order(rownames(cleaned.annotation.df)), ]

q3.log.counts.df <- q3.log.counts.df[order(colnames(q3.log.counts.df))]

# Remove .dcc from Sample ID row names
cleaned.annotation.df$sample_ID <- sub("\\.dcc", "", cleaned.annotation.df$sample_ID)

# Combine Class and Region into a combined field
cleaned.annotation.df$class_region_segment <- paste0(cleaned.annotation.df$class, "_", cleaned.annotation.df$region, "_", cleaned.annotation.df$segment)

# Create a mapping for the sample_name combination
all.samples <- c("SB-22-2963", 
 "SB-18-4726", 
 "SB-22-696", 
 "SB-22-2794", 
 "323473", 
 "TB0118", 
 "TB0133", 
 "323477", 
 "TB0144", 
 "338852", 
 "TB0164", 
 "343961")

combined.samples <- c("SB-22-2963", 
 "SB-18-4726", 
 "SB-22-696", 
 "SB-22-2794", 
 "TB0118 / 323473", 
 "TB0118 / 323473", 
 "TB0133 / 323477", 
 "TB0133 / 323477", 
 "TB0144 / 338852", 
 "TB0144 / 338852", 
 "TB0164 / 343961", 
 "TB0164 / 343961")
 
sample_mapping <- as.data.frame(list(sample_name = all.samples, 
 combined_sample_name = combined.samples))

cleaned.annotation.df <- merge(cleaned.annotation.df, sample_mapping, by = "sample_name")

# Run the PCA 
pca.stats <- prcomp(t(q3.log.counts.df), scale. = TRUE)
pca.stats.X1_2 <- as.data.frame(pca.stats$x) %>% select("PC1", "PC2")

# Column for mapping
pca.stats.X1_2$sample_ID <- rownames(pca.stats.X1_2)

# Create a combined PC and metadata df
subset.annotation <- cleaned.annotation.df %>% select(sample_ID, combined_sample_name, class_region_segment)

# Set the colors and shapes for the pca plot
color.mapping <- c("SB-22-2963" = "darkorange", 
 "SB-18-4726" = "cyan3", 
 "SB-22-696" = "hotpink", 
 "SB-22-2794" = "purple", 
 "TB0118 / 323473" = "yellow3", 
 "TB0133 / 323477" = "black", 
 "TB0144 / 338852" = "blue", 
 "TB0164 / 343961" = "brown")

shape.mapping <- c("Biopsy_Skin_tumor_LANA_pos" = 17, 
 "Biopsy_Skin_vessel_full_ROI" = 16, 
 "Biopsy_Skin_tumor_full_ROI" = 15, 
 "PDX_Skin_tumor_LANA_pos" = 11, 
 "PDX_Skin_tumor_full_ROI" = 23)

pca.annotation.df <- merge(subset.annotation, pca.stats.X1_2, by = "sample_ID")

# Rename columns for the legend
pca.annotation.df$`Type` <- pca.annotation.df$class_region_segment
pca.annotation.df$`Name` <- pca.annotation.df$combined_sample_name

# Create the scatter plot
pca.plot <- ggplot(pca.annotation.df, aes(x = PC1, 
 y = PC2, 
 color = Name, 
 shape = Type, 
 fill = Name)) +
 geom_point(size = 3, alpha = 0.8) + 
 scale_color_manual(values = color.mapping) + 
 scale_shape_manual(values = shape.mapping) + 
 scale_fill_manual(values = color.mapping) + 
  theme(
   axis.title = element_text(size = 18),
   axis.text = element_text(size = 16),
   legend.title = element_text(size = 16),
   legend.text = element_text(size = 14)
 )


print(pca.plot)

export <- FALSE
if(export){
  
  ggsave(filename = paste0(params$results.folder, "NIDAP_results/pca/PCA_Supp_Fig6E.png"), 
         pca.plot, 
         height = 10,
         width = 12)
  
}

```

#### Supp Figure 6D

```{r}


object.q3 <- q3.normalization.output$object 

 # Gather the the normalized counts
 q3.norm.counts.df <- as.data.frame(object.q3@assayData$q_norm)

# Convert counts to log2
q3.log.counts.df <- q3.norm.counts.df %>% mutate_all(~ log2(.)) %>% rename_all(~ gsub("\\.dcc", "", .))

# Remove the negative controls from the log counts
control.probes <- c("NegProbe-WTX")
q3.log.counts.df <- q3.log.counts.df[!(rownames(q3.log.counts.df) %in% control.probes), ]

# Load the annotation (same for both normalization types)
annotation <- pData(object.q3)
rownames(annotation) <- rownames(pData(object.q3))
annotation$sample_ID <- rownames(annotation)

# Remove NTCs
cleaned.annotation.df <- as.data.frame(annotation[annotation$'slide_name' != "No Template Control", ])

# Order of rownames of annotation need to match columns of count data
cleaned.annotation.df <- cleaned.annotation.df[order(rownames(cleaned.annotation.df)), ]

q3.log.counts.df <- q3.log.counts.df[order(colnames(q3.log.counts.df))]

# Remove .dcc from Sample ID row names
cleaned.annotation.df$sample_ID <- sub("\\.dcc", "", cleaned.annotation.df$sample_ID)

# Combine Class and Region into a combined field
cleaned.annotation.df$class_region_segment <- paste0(cleaned.annotation.df$class, "_", cleaned.annotation.df$region, "_", cleaned.annotation.df$segment)

# Create a mapping for the sample_name combination
all.samples <- c("SB-22-2963", 
 "SB-18-4726", 
 "SB-22-696", 
 "SB-22-2794", 
 "323473", 
 "TB0118", 
 "TB0133", 
 "323477", 
 "TB0144", 
 "338852", 
 "TB0164", 
 "343961")

combined.samples <- c("SB-22-2963", 
 "SB-18-4726", 
 "SB-22-696", 
 "SB-22-2794", 
 "TB0118 / 323473", 
 "TB0118 / 323473", 
 "TB0133 / 323477", 
 "TB0133 / 323477", 
 "TB0144 / 338852", 
 "TB0144 / 338852", 
 "TB0164 / 343961", 
 "TB0164 / 343961")
 
sample_mapping <- as.data.frame(list(sample_name = all.samples, 
 combined_sample_name = combined.samples))

cleaned.annotation.df <- merge(cleaned.annotation.df, sample_mapping, by = "sample_name")

# Run the PCA 
pca.stats <- prcomp(t(q3.log.counts.df), scale. = TRUE)
pca.stats.X1_2 <- as.data.frame(pca.stats$x) %>% select("PC1", "PC2")

# Column for mapping
pca.stats.X1_2$sample_ID <- rownames(pca.stats.X1_2)

# Create a combined PC and metadata df
subset.annotation <- cleaned.annotation.df %>% select(sample_ID, combined_sample_name, class_region_segment)

# Set the colors and shapes for the pca plot
color.mapping <- c("SB-22-2963" = "darkorange", 
 "SB-18-4726" = "cyan3", 
 "SB-22-696" = "hotpink", 
 "SB-22-2794" = "purple", 
 "TB0118 / 323473" = "yellow3", 
 "TB0133 / 323477" = "black", 
 "TB0144 / 338852" = "blue", 
 "TB0164 / 343961" = "brown")

color.mapping.type <- c("Biopsy_Skin_tumor_LANA_pos" = "orange", 
 "Biopsy_Skin_vessel_full_ROI" = "blue", 
 "Biopsy_Skin_tumor_full_ROI" = "magenta", 
 "PDX_Skin_tumor_LANA_pos" = "brown", 
 "PDX_Skin_tumor_full_ROI" = "yellow3")

shape.mapping <- c("Biopsy_Skin_tumor_LANA_pos" = 17, 
 "Biopsy_Skin_vessel_full_ROI" = 16, 
 "Biopsy_Skin_tumor_full_ROI" = 15, 
 "PDX_Skin_tumor_LANA_pos" = 11, 
 "PDX_Skin_tumor_full_ROI" = 23)

pca.annotation.df <- merge(subset.annotation, pca.stats.X1_2, by = "sample_ID")

# Rename columns for the legend
pca.annotation.df$`Type` <- pca.annotation.df$class_region_segment
pca.annotation.df$`Name` <- pca.annotation.df$combined_sample_name

# Create the scatter plot
pca.plot <- ggplot(pca.annotation.df, aes(x = PC1, 
 y = PC2, 
 color = Type, 
 fill = Type)) +
 geom_point(size = 3, alpha = 0.8) + 
 scale_color_manual(values = color.mapping.type) + 
 scale_fill_manual(values = color.mapping.type) + 
  theme(
   axis.title = element_text(size = 18),
   axis.text = element_text(size = 16),
   legend.title = element_text(size = 16),
   legend.text = element_text(size = 14)
 )


print(pca.plot)

export <- FALSE
if(export){
  
  ggsave(filename = paste0(params$results.folder, "NIDAP_results/pca/PCA_Supp_Fig6D.png"), 
         pca.plot, 
         height = 10,
         width = 12)
  
}

```

### MA Plots

```{r}
# Set up pre and post normalization counts and convert to log

pre.norm.counts <-
  log(as.data.frame(q3.normalization.output$object@assayData$exprs), base = 2)

post.q3norm.logcounts <- log(as.data.frame(q3.normalization.output$object@assayData$q_norm), base = 2)

post.negnorm.logcounts <- log(as.data.frame(neg.normalization.output$object@assayData$neg_norm), base = 2)

quantile.normalization.logcounts <- log(quantile.normalization.counts, base = 2)

# Setup annotation groups for the ratio "M" (log A - log B)
region.types <- c("Biopsy_Skin", "PDX_Skin")

# Define the name of the contrast
contrast.name <- paste0(region.types[[1]], 
                        "_", 
                        region.types[[2]])

# Set up the annotations and raw counts for the MA plots
contrast.field <- "class"
condition.label <- "Biopsy_Skin"
reference.label <- "PDX_Skin"
raw.log.counts <- pre.norm.counts
annotation.MA <- annotation
```

#### Q3 Normalization

```{r MA Plot Q3, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
# MA plots for two annotation groups to evaluate negative normalization
log.counts <- post.q3norm.logcounts

MA.plots.q3 <- make_MA(contrast.field = contrast.field, 
                       condition.label = condition.label, 
                       reference.label = reference.label, 
                       log.counts = log.counts, 
                       raw.log.counts = raw.log.counts, 
                       annotation = annotation.MA)
  
  
grid.draw(MA.plots.q3)
```

#### Negative Normalization

```{r MA Plot Neg, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}

# MA plots for two annotation groups to evaluate negative normalization
log.counts <- post.negnorm.logcounts

MA.plots.neg <- make_MA(contrast.field = contrast.field, 
                    condition.label = condition.label, 
                    reference.label = reference.label, 
                    log.counts = log.counts, 
                    raw.log.counts = raw.log.counts, 
                    annotation = annotation.MA)
  
  
grid.draw(MA.plots.neg)
```

#### Quantile Normalization

```{r MA Plot Quant, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}

# MA plots for two annotation groups to evaluate negative normalization
log.counts <- quantile.normalization.logcounts

MA.plots.quant <- make_MA(contrast.field = contrast.field, 
                    condition.label = condition.label, 
                    reference.label = reference.label, 
                    log.counts = log.counts, 
                    raw.log.counts = raw.log.counts, 
                    annotation = annotation.MA)
  
  
grid.draw(MA.plots.quant)


```

### Genes of Interest Counts

```{r Genes of Interest Counts}


CXCR4.counts <- as.data.frame(t((q3.counts["CXCR4",,drop = FALSE])))
CXCL12.counts <- as.data.frame(t((q3.counts["CXCL12",,drop = FALSE])))


# Compare to NIDAP
NIDAP.all.counts <- read.csv(paste0(params$results.folder, 
                                      "NIDAP_results/", 
                                      "GSE290108_CXCR4_updated.csv"))

NIDAP.CXCR4.counts <- NIDAP.all.counts %>% 
  filter(CXCR4 != "NA")

final.NIDAP.AOIs <- NIDAP.CXCR4.counts$Sample_ID



# What AOIs are missing in the original set from the NIDAP list 
diff.NIDAP.AOIs <- setdiff(NIDAP.all.counts$Sample_ID, gsub("\\.dcc", "", rownames(pData(init.object))))

all.annotation <- pData(init.object)
all.annotation$sample_ID <- gsub(".dcc", "", rownames(all.annotation))

annotation.diff.init_v_NIDAP.AOIs <- all.annotation %>% 
  filter(sample_ID %in% diff.NIDAP.AOIs)

# What AOIs are missing in the filtered set from the NIDAP list 

diff.NIDAP.AOIs <- setdiff(final.NIDAP.AOIs, gsub("\\.dcc", "", colnames(q3.counts)))

all.annotation <- pData(init.object)
all.annotation$sample_ID <- gsub(".dcc", "", rownames(all.annotation))

annotation.diff.filter_v_NIDAP.AOIs <- all.annotation %>% 
  filter(sample_ID %in% diff.NIDAP.AOIs)



# What are the AOIs missing from the NIDAP list but in the original set
diff.original.AOIs <- setdiff(gsub("\\.dcc", "", colnames(q3.counts)), final.NIDAP.AOIs)

annotation.original.AOIs <- all.annotation %>% 
  filter(sample_ID %in% diff.original.AOIs)

# Create the CXCR4 and CXCL12 data frame with annotations
q3.counts.subset <- q3.counts %>% 
  filter(gene %in% c("CXCR4", "CXCL12"))

# Format for merging with annotation
counts.t <- as.data.frame(t(q3.counts.subset))
counts.t <- counts.t[2:length(counts.t$CXCL12),] # remove first row with gene names
counts.t$segmentID <- rownames(counts.t)

counts.anno.df <- merge(cleaned.annotation.df, counts.t, by = "segmentID")

counts.anno.df.final <- counts.anno.df %>% 
  select(sample_ID, 
         slide_name, 
         segment, 
         class, 
         region, 
         Tumor, 
         Source, 
         LANA_segmented, 
         roi, 
         CXCR4, 
         CXCL12)

export <- FALSE
if(export){
  
  write.csv(counts.anno.df.final, 
            file = paste0(params$project.folder, 
                          "NIDAP_Workbook_Restoration/", 
                          "CXCR4_CXCL12_counts_annotation.csv"),  
            row.names = FALSE)
  
}
```
